Type theories as quotient inductive-recursive types

### 2025-02-05 tex/main.tex
Draft S4.1

\subsection{... with an empty universe} \label{subsec:SC+U}
\LT[noinline]{Shall we go for Coquand universes? \cite{Coquand2013}}

In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension adds following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}}
\end{alignat*}
as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.

@@

  For other two equality constructors, the arguments follow similarly.
\end{proof}
The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
\begin{proposition}[Coherence, II]
  For the following equality constructors $p : \sigma = \tau$
  \begin{alignat*}{5}
    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
  \end{alignat*}
   the identities
  \danger
  \begin{align*}
  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
%    &
%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
  \end{align*}
  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
\end{proposition}
\begin{proof}
  We demonstrate the proof for $\pi_1\beta: \pi_1(\sigma, t) = t$ and other three cases follow similarly.

  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$, assuming the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$, which are then proved by induction on $\Xi$.
  \begin{enumerate}
    \item By induction on $A$, consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$.
      \begin{description}
        \item[$A$ is $\UU$:] By definition
          \[
            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
          \]
          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by the congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
          \LT[noinline]{I think the congruence for $\UU$ is true but not sure how precisely.}

@@

        \item[$A$ is $\Pi\,B\,C$:]
          Note that one of the induction hypothesises is
          \[
            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,C = [\sigma \upuparrows \Xi]_{\Ty}\,C
          \]
          for \emph{arbitrary} $\Xi$, so we can apply the induction hypothesis to $B$ and $C$ below:
          \begin{align*}
            & [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) \\
            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$} \\
            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows (\Xi, B)]\,C) && \text{by the definition of $\upuparrows$} \\
            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by the induction hypothesises} \\
            \judgeq{} & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([(\sigma \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $\upuparrows$} \\
            \judgeq{} & [\sigma\upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$.}
          \end{align*}
          Note that 
      \end{description}
    \item By induction on $\Xi$, consider following cases for $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [\sigma]_{\Tel}\,\Xi$. 
      \begin{description}
        \item[$\Xi$ is the empty telescope $\emptytel$:] By definition, $[\pi_1(\sigma, t)]_{\Tel}\,\emptytel \judgeq \emptytel \judgeq [\sigma]_{\Tel}\,\emptytel$.
        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
          \begin{align*}
            [\pi_1(\sigma, t)]_{\Tel}\,(\Xi, A) & \judgeq [\pi_1(\sigma, t)]_{\Tel}\,\Xi, [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,A && \text{by definition} \\
                                                & = [\sigma]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
                                                & \judgeq [\sigma]_{\Tel}(\Xi, A) && \text{by definition}
          \end{align*}
      \end{description}
    \item By induction on $\Xi$, consider the following cases for $\pi_1(\sigma, t) \upuparrows \Xi = \sigma \upuparrows\,\Xi$.
      \begin{description}
        \item[$\Xi$ is the empty telescope $\emptytel$:] $\pi_1(\sigma, t) \upuparrows \emptytel \judgeq \pi_1(\sigma, t) = \sigma \judgeq \sigma \upuparrows \emptytel$ by definition and the equality constructor.
        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
          \begin{align*}
            \pi_1(\sigma, t) \upuparrows (\Xi, A) & \judgeq (\pi_1(\sigma, t) \upuparrows \Xi) \uparrow A \\
                                                  & =^{} (\pi_1(\sigma, t) \upuparrows \Xi)^{+} \\
                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus[\gamma]_{\Tel}\,\Xi, [\gamma \upuparrows \Xi]_{\Ty}\,A)\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
                                                  & = (\sigma \upuparrows \Xi) \uparrow A \\
                                                  & \judgeq \sigma \upuparrows (\Xi , A)
          \end{align*}
      \end{description}
  \end{enumerate}

@@

\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the unit type.
  \item Discuss the Boolean type.
  \item Discuss the type of natural numbers.
  \item Discuss the extensional identity type.
\end{enumerate}
}

\section{The eliminator, the standard model, and the equivalence to the QII definition of type theory}
\LT{%
\begin{enumerate}
  \item Introduce motives and methods for the definition in \Cref{subsec:SC+U+Pi}.
  \item The rule $\mathsf{Elim}_{\Ty}\,([ \sigma ]_{\Ty}\,A) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Ty}(\mathsf{Elim}_{\Ty}\,A)$ is not defined but provable. 
  \item $\mathsf{Elim}_{\Tm}\,([ \sigma ]_{\Tm}\,t) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Tm}(\mathsf{Elim}_{\Tm}\,t)$ is proved.
\end{enumerate}
}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
After giving the QIIR definition of a type theory, we have to define how to \emph{eliminate} types $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ altogether to another types (called \emph{motives}) by functions (called \emph{methods}) subject to equations corresponding to point constructors and equality constructors (\cref{subsec:eliminator}).
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart (\cref{subsec:equivalence}).

\subsection{Eliminator} \label{subsec:eliminator}
As the same as the QII definition~\cite{Altenkirch2016a}, the motives for the QIIR definition are
\begin{align*}
  \Ctx^A & : \Ctx \to \Set &
  \Sub^A & : \implicit{\Gamma, \Delta}\,\Ctx^A\;\Gamma \to \Ctx^A\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^A  & : \implicit{\Gamma}\,\Ctx^A\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^A  & : \implicit{\Gamma, A}\; (\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
and each method corresponds to each constructor, e.g., 
\[
  \emptyctx^A : \Ctx^A\;\emptyctx 
  \quad
  \blank ,^A \blank : \implicit{\Gamma, A}\;(\Gamma^A : \Ctx^A\;\Gamma) \to \Ty^A\;\Gamma^A\; A \to \Ctx^A\;(\Gamma, A)
\]
\subsection{The standard model} \label{subsec:std-model}
\subsection{The equivalence} \label{subsec:equivalence}
\section{Formal implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

>>>>

\subsection{... with an empty universe} \label{subsec:SC+U}
In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension adds following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}}
\end{alignat*}
as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.

@@

  For other two equality constructors, the arguments follow similarly.
\end{proof}
The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
\begin{proposition}[Coherence, II]
  For the following equality constructors $p : \sigma = \tau$
  \begin{alignat*}{5}
    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
  \end{alignat*}
   the identities
  \danger
  \begin{align*}
  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
%    &
%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
  \end{align*}
  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
\end{proposition}
\begin{proof}
  We show the statement for $\pi_1\beta: \pi_1(\sigma, t) = t$ and the other three cases follow similarly.

  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$ and the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$ by induction on $\Xi$.
  \begin{enumerate}
    \item Consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$ by induction on $A$.
      \begin{description}
        \item[$A$ is $\UU$:] By definition
          \[
            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
          \]
          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
          \LT[noinline]{To fix: Clarify this.}

@@

        \item[$A$ is $\Pi\,B\,C$:]
          Note that one of the induction hypothesises is
          \[
            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,C = [\sigma \upuparrows \Xi]_{\Ty}\,C
          \]
          for \emph{arbitrary} $\Xi$, so we can apply the induction hypothesis to $B$ and $C$ below:
          \begin{align*}
            & [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) \\
            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$} \\
            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows (\Xi, B)]\,C) && \text{by the definition of $\upuparrows$} \\
            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by induction hypothesises} \\
            \judgeq{} & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([(\sigma \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $\upuparrows$} \\
            \judgeq{} & [\sigma\upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$.}
          \end{align*}
      \end{description}
    \item By induction on $\Xi$, consider following cases for $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [\sigma]_{\Tel}\,\Xi$. 
      \begin{description}
        \item[$\Xi$ is the empty telescope $\emptytel$:] By definition, $[\pi_1(\sigma, t)]_{\Tel}\,\emptytel \judgeq \emptytel \judgeq [\sigma]_{\Tel}\,\emptytel$.
        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
          \begin{align*}
            [\pi_1(\sigma, t)]_{\Tel}\,(\Xi, A) & \judgeq [\pi_1(\sigma, t)]_{\Tel}\,\Xi, [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,A && \text{by definition} \\
                                                & = [\sigma]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
                                                & \judgeq [\sigma]_{\Tel}(\Xi, A) && \text{by definition}
          \end{align*}
      \end{description}
    \item By induction on $\Xi$, consider the following cases for $\pi_1(\sigma, t) \upuparrows \Xi = \sigma \upuparrows\,\Xi$.
      \begin{description}
        \item[$\Xi$ is the empty telescope $\emptytel$:] $\pi_1(\sigma, t) \upuparrows \emptytel \judgeq \pi_1(\sigma, t) = \sigma \judgeq \sigma \upuparrows \emptytel$ by definition and the equality constructor.
        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
          \begin{align*}
            \pi_1(\sigma, t) \upuparrows (\Xi, A) & \judgeq (\pi_1(\sigma, t) \upuparrows \Xi) \uparrow A \\
                                                  & =^{} (\pi_1(\sigma, t) \upuparrows \Xi)^{+} \\
                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus[\gamma]_{\Tel}\,\Xi, [\gamma \upuparrows \Xi]_{\Ty}\,A)\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
                                                  & = (\sigma \upuparrows \Xi) \uparrow A \\
                                                  & \judgeq \sigma \upuparrows (\Xi , A)
          \end{align*}
      \end{description}
  \end{enumerate}

@@

\subsection{... and other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the extensional identity type.
  \item Discuss Coquand universes.
  \item Discuss the unit type.
  \item Discuss the Boolean type.
  \item Discuss the type of natural numbers.
\end{enumerate}
}

\section{Elimination and models}
\LT{%
\begin{enumerate}
  \item Introduce motives and methods for the definition in \Cref{subsec:SC+U+Pi}.
  \item The rule $\mathsf{Elim}_{\Ty}\,([ \sigma ]_{\Ty}\,A) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Ty}(\mathsf{Elim}_{\Ty}\,A)$ is not defined but provable. 
  \item $\mathsf{Elim}_{\Tm}\,([ \sigma ]_{\Tm}\,t) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Tm}(\mathsf{Elim}_{\Tm}\,t)$ is proved.
\end{enumerate}
}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types (called \emph{motives}) by functions (called \emph{methods}) subject to equations corresponding to point constructors and equality constructors (\cref{subsec:eliminator}).
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart (\cref{subsec:equivalence}).

\subsection{The elimination principle} \label{subsec:eliminator}
To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify four types
\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
called the \emph{motivates} and functions indexed by corresponding constructors such as 
\begin{alignat*}{3}
  \emptyctx^M     & :                      && \Ctx^M\;\emptyctx \\
  \blank,^M\blank & : \implicit{\Gamma, A} && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
  \idS^M          & : \implicit{\Gamma, \Gamma^M}    && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
  \blank;\blank   & : \implicit{\Gamma, \Gamma^M, \Delta, \Delta^M, \Theta, \Theta^M} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Sub^M\;\Delta^M\;\Theta^M\;\tau \to \Sub^M\;\Gamma^M\;\Theta^M\;(\sigma;\tau)\\
  [\blank]^M\blank & : \implicit{\Gamma, \Gamma^M, \Delta, \Delta^M, \sigma, A, A^M, t} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \\
  &&& \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
\end{alignat*}
as well as recursive functions
\begin{alignat*}{3}
  [\blank]_{\Ty}^M\blank & : \implicit{\Gamma, \Gamma^M, \Delta, \Delta^M, \sigma, A} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
  [\blank]_{\Tm}^M\blank & : \implicit{\Gamma, \Gamma^M, \Delta, \Delta^M, \sigma, A, A^M, t} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \\
                         &&& \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
\end{alignat*}
called \emph{methods} subject to the corresponding equations \emph{and function clauses}, e.g., 
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\dots} && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
  [;]_{\Ty}^M & : \implicit{\dots} && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
\end{alignat*}
To define the elimination principle, we cannot avoid the use of transports in identities that would need transports in its QII definition of syntax, such as $[\idS]_{\Tm}^M$, since identities for substitutions (e.g., $[id]_{\Ty}^M$) are still propositional. 
On the other hand, in the elimination principle for the QII definition, the above identity $[\idS]_{\Ty}^M$ already requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the type of $A^M$.
For brevity, we omit the rest of methods and we refer to the Agda implementation for the complete definition of motives and methods.
\LT[noinline]{Add a link.}

Given the motives and the methods, the elimination principle is defined as functions
\begin{alignat*}{5}
  \elim_{\Ctx} & : (\Gamma : \Ctx) \to \Ctx^M\;\Gamma \\
  \elim_{\Ty}  & : (A : \Ty\;\Gamma) \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \\
  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
  \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t
\end{alignat*}
by interpreting each constructor using the corresponding method inductively.

In the QII counterpart, $[\blank]\blank$ is a constructor, so $\elim_{\Ty}\;([\sigma]\,A)$ is eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$ definitionally and likewise for the term substitution.
In our QIIR definition, however, these become propositional identities:
\begin{proposition}
  Given the motives and the methods, we have
\begin{alignat*}{5}
  \elim_{\Ty}[] & : \implicit{\Gamma, \Delta, \sigma, A} & \elim_{\Ty}\,([\sigma]_{\Ty}\,A) & = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
  \elim_{\Tm}[] & : \implicit{\Gamma, \Delta, \sigma, A, t} & \elim_{\Tm}\,([\sigma]_{\Tm}\,t) & =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t)
\end{alignat*}

\end{proposition}
These identities are required to interpret terms involving definitional identities.
For example, consider the case for eliminating an extension $(\sigma, t)$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$:
\[
  \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,^M \; t^M
\]
The eliminated type of $t^M$ is $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$ instead of the eliminated type $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$ of $\elim_{\Tm}\;t$, so $t^M$ above have to be the transported term
\[
  \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
\]
of $\elim_{\Tm}\,t$ along the identity $\elim_{\Ty}[]$. 
\LT{Add some remark regarding the proof}

\subsection{The standard model} \label{subsec:std-model}
\subsection{The equivalence} \label{subsec:equivalence}
\section{Formal implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}
